### 2024-04-08 chap_RL/chap_RL.tex
Chap RL - Reward and success

\section{Learning process and convergence}
\label{sec:RL:learning}

@@

To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase of the agent has been done on the monthly model. Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus about 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).

The learning phase has been split by batch of 500 steps. For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts. This makes sure replay buffer is full enough for useful updates. At the end of each batch, the up-to-date policy, \ie the \gls{NN}, is saved. This way, we can assess the progress in the learning process and its convergence (see Figure \ref{fig:reward_success}).

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Mean_reward.pdf}
\includegraphics[width=0.49\textwidth]{Success_rate.pdf}
\caption{Mean reward and success rate of the different learning batches. The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view. The success rate also confirms this progression and convergence.}
\label{fig:reward_success}
\end{figure}

\begin{itemize}
\item Convergence of reward or success rate
\item Actions to take, which ones are binding
\item Show the state space: share of \gls{VRES} and efficiency
\item Assess the severity of learning on Monthly model, versus Hourly model. Compare the policies: one versus the other versus one followed by the other
\end{itemize}

>>>>

\section{Convergence and learning process}
\label{sec:RL:learning}

@@

To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase of the agent has been done on the monthly model. Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}). \\

\myparagraph{Reward and success}\\

The learning phase has been split into batches of 500 steps. For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts. This makes sure replay buffer is full enough for useful updates. At the end of each batch, the up-to-date policy, \ie the \gls{NN}, is saved. This way, we can assess the progress in the learning process and its convergence (see Figure \ref{fig:reward_success}).  The mean reward increases rapidly at the beginning of the learning process before reaching a plateau where the optimisation of the policy becomes more marginal. Right hand side of Figure \ref{fig:reward_success} shows the success rate as the share of transitions meeting the \ce{CO2}-budget (see Section \ref{sec:cs:CO2-budget}) over the total number of attempted transitions, \ie episodes. Even though this success rate is not what drives the agent's optimisation, it shows that the shape of the reward (see Figure \ref{fig:Reward}) leads towards more and more successes.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Mean_reward.pdf}
\includegraphics[width=0.49\textwidth]{Success_rate.pdf}
\caption{Mean reward and success rate of the different learning batches. The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view. The evolution of the success rate also shows the shape of the reward aims at more and more successful transitions.}
\label{fig:reward_success}
\end{figure}

Over the entire learning process, 23\% of the episodes have been successful versus 73\% of failures, the rest being infeasible problems where the actions taken by the agent are too constraining for the model (see Section \ref{sec:RL:act_states_rew}). Besides these infeasible cases, there is a range in the reward where failures and successes overlap (see Figure \ref{fig:reward_status}). This area corresponds to either transitions that exceed the \ce{CO2}-budget in 2050 but are cheaper than the total transition cost of reference (see Section \ref{sec:RL:act_states_rew}) or successful transitions that are more expensive. Besides this overlap, we observe that successes account for the majority of the cases where the reward is positive. This is another indication that the shape of the reward is appropriate in this exploration of transition pathways.

Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year''. Beyond this point, the chances to succeed the transition are 47\%. In other words, mid-term actions are necessary to hope succeeding the transition.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Reward_status.pdf}
\includegraphics[width=0.49\textwidth]{Reward_status_2.pdf}
\caption{Reward distribution between successes and failures. Right hand side details at the end of which time-window the failure occurred. The ``tipping year'' is 2040 as failing the transition by 2040 represents 58\% of all the failures. Beyond this point, succeeding the transition represents 47\% of the cases.}
\label{fig:reward_status}
\end{figure}

\myparagraph{States: Renewables and efficiency}\\

\begin{itemize}
\item Actions to take, which ones are binding
\item Show the state space: share of \gls{VRES} and efficiency
\item Assess the severity of learning on Monthly model, versus Hourly model. Compare the policies: one versus the other versus one followed by the other
\end{itemize}
